---
title: Publications
site: distill::distill_website
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Learn more about creating websites with Distill at:
# https://rstudio.github.io/distill/website.html

```

## Peer-reviewed publications

```{r}
library(magrittr)

bibfile <- tempfile(fileext = ".bib")

httr2::request("https://api.archives-ouvertes.fr/search/?q=authIdHal_s:hugo-gruson&wt=bibtex") %>%
  httr2::req_perform() %>%
  httr2::resp_body_string() %>%
  write(bibfile)
```

```{r, results='asis'}
b <- tools::bibstyle(
  "by_year", 
  fmtPrefix = function(paper) "-", 
  .init = TRUE, 
  sortKeys = function(refs) rank(purrr::map_chr(unclass(refs), "year"))
)

bibtex::read.bib(bibfile) %>%
  purrr::keep(~ attr(unclass(.x)[[1]], "bibtype") == "Article") %>%
  sort(.bibstyle = "by_year", decreasing = TRUE) %>%
  purrr::map_chr(format, "html", .bibstyle = "by_year") %>%
  purrr::map_chr(~ gsub("Gruson H", "**Gruson H**", .x)) %>%
  paste(collapse = "\n") %>%
  cat()
```

## Invited blog posts and news items

<p>
<ul>
<li><em>Covidpreprints.com: Automating Website Updates with the europepmc and rAltmetric Packages</em> (2020), featured on <a href="https://ropensci.org/blog/2020/10/13/covidpreprints/">rOpenSci blog</a> and <a href="https://prelights.biologists.com/news/covidpreprints-com-automating-website-updates-with-the-europepmc-and-raltmetric-packages/">preLights blog</a></li>
<li><em>Community Captioning of rOpenSci Community Calls</em> (2020), featured on <a href="https://ropensci.org/blog/2020/07/15/subtitles/">rOpenSci blog</a></li>
<li><em>Access Publisher Copyright &amp; Self-Archiving Policies via the 'SHERPA/RoMEO' API</em> (2019), featured on <a href="https://ropensci.org/blog/2019/06/04/rromeo/">rOpenSci blog</a></li>
<li><em>When Standards Go Wild - Software Review for a Manuscript</em> (2019), featured on <a href="https://ropensci.org/blog/2019/04/18/wild-standards/">rOpenSci blog</a> and <a href="https://methodsblog.com/2019/04/18/ropensci-code-review-guidelines/">Methods in Ecology and Evolution blog</a></li>
<li><em>Observatoire des oiseaux des jardins : premiers résultats</em> (2014), in L'Oiseau Mag (trimestrial magazine of BirdLife France)</li>
</ul>
</p>
