---
title: Publications
site: distill::distill_website
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Learn more about creating websites with Distill at:
# https://rstudio.github.io/distill/website.html

```

## Peer-reviewed publications

*(last automatic update: `r Sys.Date()`)*

```{r}
library(magrittr)

bibfile <- tempfile(fileext = ".bib")

httr2::request("https://api.archives-ouvertes.fr/search/?q=authIdHal_s:hugo-gruson&wt=bibtex") %>%
  httr2::req_perform() %>%
  httr2::resp_body_string() %>%
  write(bibfile)
```

```{r, results='asis'}
b <- tools::bibstyle(
  "by_year", 
  fmtPrefix = function(paper) "-", 
  .init = TRUE, 
  sortKeys = function(refs) rank(purrr::map_chr(unclass(refs), "year"))
)

bibtex::read.bib(bibfile) %>%
  purrr::keep(~ attr(unclass(.x)[[1]], "bibtype") == "Article") %>%
  sort(.bibstyle = "by_year", decreasing = TRUE) %>%
  purrr::map_chr(format, "html", .bibstyle = "by_year") %>%
  purrr::map_chr(~ gsub("Gruson H", "**Gruson H**", .x)) %>%
  paste(collapse = "\n") %>%
  cat()
```

## Invited blog posts and news items

-   *System dependencies in R packages and automatic testing* (2023), featured
    on the [R-hub blog](https://blog.r-hub.io/2023/09/26/system-dependency/)

    -   Commented in the [R Weekly Highlights
        podcast](https://podverse.fm/clip/k8RPsYA1Ee)

-   *From hidden to visisble: how can we recognise research software?* (2023),
    featured on the [Software Sustainability Institute
    blog](https://software.ac.uk/blog/2023-08-01-hidden-visible-how-can-we-recognise-research-software)

-   *Ensuring & Showcasing the Statistical Correctness of your R Package*
    (2023), featured on the [Epiverse-TRACE
    blog](https://epiverse-trace.github.io/posts/statistical-correctness/)

    -   Commented in the [R Weekly Highlights
        podcast](https://podverse.fm/clip/EkvfvBF6AW)

-   *Minimum R version dependency* (2022), featured on the [R-hub
    blog](https://blog.r-hub.io/2022/09/12/r-dependency/)

-   *Opening the door to new contributors in open source projects* (2022),
    featured on the [Software Sustainability Institute
    blog](https://www.software.ac.uk/blog/2022-06-22-opening-door-new-contributors-open-source-projects).

-   *Why You Should (or Shouldn't) Build an API client?* (2022), featured on
    [rOpenSci
    blog](https://ropensci.org/blog/2022/06/16/publicize-api-client-yes-no/).

    -   Commented in the [R Weekly Highlights
        podcast](https://share.fireside.fm/episode/87RSVeFz+4qaUu8e0?t=538)

-   *Dealing with flaky GitHub Actions* (2022), featured on [EpiForecasts
    blog](https://epiforecasts.io/posts/2022-04-11-robust-actions/robust-actions.html).

-   *Checking the Inputs of your R Functions* (2022), featured on [R-hub
    blog](https://blog.r-hub.io/2022/03/10/input-checking/).

-   *Covidpreprints.com: Automating Website Updates with the europepmc and
    rAltmetric Packages* (2020), featured on [rOpenSci
    blog](https://ropensci.org/blog/2020/10/13/covidpreprints/) [preLights
    blog](https://prelights.biologists.com/news/covidpreprints-com-automating-website-updates-with-the-europepmc-and-raltmetric-packages/).

-   *Community Captioning of rOpenSci Community Calls* (2020), featured on
    [rOpenSci blog](https://ropensci.org/blog/2020/07/15/subtitles/).

-   *Access Publisher Copyright & Self-Archiving Policies via the 'SHERPA/RoMEO'
    API* (2019), featured on [rOpenSci
    blog](https://ropensci.org/blog/2019/06/04/rromeo/).

-   *When Standards Go Wild - Software Review for a Manuscript* (2019), featured
    on [rOpenSci blog](https://ropensci.org/blog/2019/04/18/wild-standards/) and
    [Methods in Ecology and Evolution
    blog](https://methodsblog.com/2019/04/18/ropensci-code-review-guidelines/).

-   *Observatoire des oiseaux des jardins : premiers résultats* (2014), in
    L'Oiseau Mag (trimestrial magazine of BirdLife France).
