---
title: "Static Analysis in R and for Bioconductor"
logo: "EMBL_logo_colour_DIGITAL.png"
footer: "Hugo Gruson; Bioconductor TAB meeting 02/2026"
subtitle: "Bioconductor TAB meeting 02/2026"
date: "2026-02-05"
author: "Hugo Gruson"
format: 
  clean-revealjs:
    slide-number: true
---

## Definition

Static code analysis and linting refer to the concept of detecting incorrect, suboptimal or dangerous code patterns **without executing the code**, only by analysing the code syntax tree.

## Rationale

Static analysis:

- may catch bugs early
- may improve code readability, maintainability, performance

## Benefits over testing

- fast: even for packages doing heavy computations
- safe: untrusted, potentially unsafe code, doesn't need to be executed
- robust: non-reproducible code or incomplete code (e.g., missing data) can be analyzed as long as it's valid code. 

## Static analysis frameworks in the R ecosystem

| Tool / framework         | Runs in Bioconductor | Runs in CRAN | Runs in rOpenSci |
|--------------------------|----------------------|--------------|------------------|
| R CMD check              | &#x2714;             | &#x2714;     | &#x2714;         |
| biocCheck                | &#x2714;             | &#x274C;     | &#x274C;         |
| rchk                     | &#x274C;             | &#x2714;     | &#x274C;         |
| lintr ( / flir / jarl )  | &#x274C;             | &#x274C;     | (&#x2714; *)     |

*: during initial onboarding only 

## biocCheck

Focused on Bioconductor best practices.

Well-known to this group.

## lintr

Not just style!

. . .

Categories

```{r}
lintr::available_tags("lintr")
```

Examples

```{r}
lintr::lint(
  text = "length(x == 0)",
  linters = lintr::length_test_linter()
)
lintr::lint(
  text = "x == NA",
  linters = lintr::equals_na_linter()
)
```

## lintr vs biocCheck {.smaller}

:::: {.columns}

::: {.column width="70%"}

| biocCheck check                                      | linter                                |
|------------------------------------------------------|---------------------------------------|
| `vapply()` instead of `sapply()`                     | `undesirable_function_linter()`       |
| `seq_len()` or `seq_along()` instead of `1:length()` | `seq_linter()`                        |
| `pkg:fun()` instead of `pkg::fun()`                  | No                                    |
| Download links to GitHub/GitLab/BitBucket            | No                                    |
| `download.file()` in `onLoad()`                      | No                                    |
| `paste()` in condition messages                      | `condition_message_linter()`          |
| `message()` instead of `cat()`/`print()`             | `print_linter()`                      |
| `<-` instead of `=`                                  | `assignment_linter()`                 |
| No `.Deprecated`/`.Defunct`/etc.                     | No                                    |
| `TRUE`/`FALSE` instead of `T`/`F`                    | `T_and_F_symbol_linter()`             |
| `is()` instead of `class() ==`                       | `class_equals_linter()`               |
| `sytem2()` instead of `system()`                     | No                                    |
| No `set.seed()`                                      | No                                    |
| No direct slot access (`@`)                          | No                                    |
| No `browser()`                                       | `undesirable_function_linter()`       |
| No `<<-`                                             | `undesirable_operator_linter()`       |
| No `Sys.setenv()`                                    | `undesirable_function_linter()`       |
| No `suppressWarnings()`                              | No                                    |

:::

::: {.column width="30%"}

All missing linters could very easily be implemented!

:::

::::


## lintr vs biocCheck proposal

- Opportunity: Cross-pollination between biocCheck and lintr:
  - reduce maintenance burden
  - increase user base (= more test cases, more feedback)
  - make Bioconductor checks available to the wider R community

- Proposal: bioconductor category

```{.r}
lintr::lint_package(linters = lintr::linters_with_tags("bioconductor"))
```

## R CMD check vs biocCheck

Some shared checks, such as:

```{.r}
class(x) == "myClass"
```

Not necessarily an issue(?)

## rchk: static analysis for C code in R packages

`PROTECT()` / `UNPROTECT()` balance and common mistakes in C code interacting with the R C API.

Additional resources:

- [rchk source and documentation](https://github.com/kalibera/rchk)
- [Blog post: "Common PROTECT Errors", by Tomas Kalibera](https://developer.r-project.org/Blog/public/2019/04/18/common-protect-errors/)

## rchk in CRAN & Bioconductor

- Routinely runs on CRAN packages
- [Ran on Bioconductor packages in 2017](https://stat.ethz.ch/pipermail/bioc-devel/2017-April/010771.html)
   - Some packages were not fixed until recently / are not fixed yet
   - No information about newer releases

 Value for our community:

- Available to individual developers as a Docker image or as a GitHub Action via R-hub
  - May be hard to set up
  - Upstream Docker image not compatible with latest Bioconductor releases
- Bioconductor has a lot of native C code

## Proposal 1: new checks on Bioconductor

| Tool / framework         | Runs in Bioconductor | Runs in CRAN | Runs in rOpenSci |
|--------------------------|----------------------|--------------|------------------|
| R CMD check              | &#x2714;             | &#x2714;     | &#x2714;         |
| biocCheck                | &#x2714;             | &#x274C;     | &#x274C;         |
| rchk                     | 💡                   | &#x2714;     | &#x274C;         |
| lintr ( / flir / jarl )  | 💡 (via biocCheck)   | &#x274C;     | (&#x2714; *)     |

New runners for checks could be hosted at EMBL, since:

- proposed checks are entirely independent from BBS
- we already have a full copy of the sources

## Proposal 2: integration of lintr in biocCheck

Share some code components between biocCheck and lintr:

- reduce maintenance burden
- increase user base (= more test cases, more feedback)
- make Bioconductor checks available to the wider R community

## Discussion
