---
title: "How to choose a license for an academic project"
author: "Hugo Gruson"
format: revealjs
date: ""
---

## Free/Libre Open Source Software ("FLOSS")

Four freedoms: 

- The freedom to run the program as you wish, for any purpose.
- The freedom to study how the program works, and change it so it does your computing as you wish.
- The freedom to redistribute copies so you can help others.
- The freedom to distribute copies of your modified versions to others. By doing this you can give the whole community a chance to benefit from your changes.

.footnote[Source: <https://www.gnu.org/philosophy/free-sw.en.html>]

---

## Copyleft licenses

You have to keep the same license (or a compatible FLOSS license) for downstream
projects.

- GNU Public License (version 2 & 3) and derivatives:
  - AGPL (for applications on servers mainly)
  - LGPL (for libraries mainly)

. . .

- Creative Commons ShareAlike ("CC-SA")

---

## Copyleft licenses

### GNU Public License notable examples

{{< fa brands linux size=4x >}}
{{< fa brands git size=4x >}}
{{< fa brands r-project size=4x >}}

. . .

### Creative Commons ShareAlike notable examples

{{< fa brands wikipedia-w size=4x >}}
{{< fa brands stack-overflow size=4x >}}

---

## Non-copyleft licenses

- MIT
- Apache
- BSD
- CC0 / zero-clause BSD / public domain

---

## Non-copyleft licenses

### MIT notable examples

{{< fa brands react size=2x >}}
{{< fa brands angular size=2x >}}
{{< fa brands rust size=2x >}}

### BSD(-like) notable examples

{{< fa brands python size=2x >}}
numpy, pandas, scipy, pytorch

### Apache notable examples

{{< fa brands android size=2x >}}

---

## Other notable cases {.smaller}

Intermediary copyleft:

- MPL: only redistribute parts under copyleft, not the whole codebase.

Relicensing clause:

- Artistic 2.0: the copyleft project can adopt another copyleft license.

. . .

Non-free licenses:

- Creative Commons NoDerivative / NoCommercial (CC-ND / CC-NC)

. . .

- JSON license

> The Software shall be used for Good, not Evil.

---

## A resource to help you choose

[https://choosealicense.com](https://choosealicense.com){preview-link="true"}

---

## Licenses in CRAN

As of `r Sys.Date()`:

```{r, cache = TRUE, cache.extra = Sys.Date()}
cran_today <- available.packages(contriburl = contrib.url("https://cloud.r-project.org"))
```

```{r, figure = TRUE}
library(dplyr)
library(ggplot2)
cran_today %>%
  as.data.frame() %>%
  mutate(simple_license = case_when(
    grepl("^AGPL", License) ~ "AGPL",
    grepl("^GNU Affero", License) ~ "AGPL",
    grepl("^GPL", License) ~ "GPL",
    grepl("^GNU General Public License", License) ~ "GPL",
    grepl("^LGPL", License) ~ "LGPL",
    grepl("^Lesser G[nN][uU]", License) ~ "LGPL",
    grepl("GNU Lesser", License) ~ "LGPL",
    grepl("^MIT", License) ~ "MIT",
    grepl("^MPL", License) ~ "MPL",
    grepl("^Mozilla", License) ~ "MPL",
    grepl("^Apache License", License) ~ "Apache",
    grepl("^BSD(_| )2", License) ~ "BSD 2-clause",
    grepl("^BSD(_| )3", License) ~ "BSD 3-clause",
    grepl("^Artistic", License) ~ "Artistic",
    grepl("^CC0", License) ~ "Public domain",
    grepl("Unlimited", License) ~ "Public domain",
    grepl("^CC", License) ~ "Creative Commons",
    grepl("^Creative Commons", License) ~ "Creative Commons",
    TRUE ~ "Other"
  )) %>%
  count(simple_license) %>%
  ggplot(aes(y = reorder(simple_license, n), x = n)) +
    geom_histogram(stat = "identity") +
    labs(y = "Licence family", x = "Number of packages") +
    theme_minimal()
```

## Licenses in Bioconductor

As of `r Sys.Date()`:

```{r, cache = TRUE, cache.extra = Sys.Date()}
bioc_today <- available.packages(contriburl = contrib.url("https://bioconductor.org/packages/release/bioc"))
```

```{r}
library(dplyr)
library(ggplot2)
bioc_today %>%
  as.data.frame() %>%
  mutate(simple_license = case_when(
    grepl("^AGPL", License) ~ "AGPL",
    grepl("^GNU Affero", License) ~ "AGPL",
    grepl("^GPL", License) ~ "GPL",
    grepl("^GNU General Public License", License) ~ "GPL",
    grepl("^LGPL", License) ~ "LGPL",
    grepl("^Lesser G[nN][uU]", License) ~ "LGPL",
    grepl("GNU Lesser", License) ~ "LGPL",
    grepl("^MIT", License) ~ "MIT",
    grepl("^MPL", License) ~ "MPL",
    grepl("^Mozilla", License) ~ "MPL",
    grepl("^Apache License", License) ~ "Apache",
    grepl("^BSD(_| )2", License) ~ "BSD 2-clause",
    grepl("^BSD(_| )3", License) ~ "BSD 3-clause",
    grepl("^Artistic", License) ~ "Artistic",
    grepl("^CC0", License) ~ "Public domain",
    grepl("Unlimited", License) ~ "Public domain",
    grepl("^CC", License) ~ "Creative Commons",
    grepl("^Creative Commons", License) ~ "Creative Commons",
    TRUE ~ "Other"
  )) %>%
  count(simple_license) %>%
  ggplot(aes(y = reorder(simple_license, n), x = n)) +
    geom_histogram(stat = "identity") +
    labs(y = "Licence family", x = "Number of packages") +
    theme_minimal()
```