---
title: "Compare S4 standardGeneric() special case to non-special case"
draft: true
format: html
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

S4 style guides usually recommend not adding braces around `standardGeneric()` function bodies.
For examples, the ['S4' chapter of 'Advanced R'](https://adv-r.hadley.nz/s4.html#s4-generics) states:

> It is bad practice to use {} in the generic as it triggers a special case that is more expensive, and generally best avoided.
>
> ```r
> # Don't do this!
> setGeneric("myGeneric", function(x) {
>   standardGeneric("myGeneric")
> })
> ```

This is a common source of confusion for novice S4 users, and for tools analyzing or modifying R code formatting (https://github.com/posit-dev/air/issues/398, https://github.com/r-lib/lintr/issues/1968, https://github.com/hadley/adv-r/issues/1804)
Indeed, it is unusual in R to have formatting choices result in different behaviours.
Furthermore, the exact consequence of the special case seem to have been lost in time and remains unknown to many R users, including experienced S4 developers.
It is unclear if the recommendation is still necessary today, or if we are just following a historical convention.

This blog post explores where this special case arises from, and what its measurable consequences (if any) are.

## Special case definition

The source of the special case was pointed out to me by Lionel Henry on the Air bug tracker. [In the methods package, there is an extra check that will compare the body of our defined generic to a known pattern for `standardGeneric()` calls without braces](https://github.com/r-devel/r-svn/blob/097aac926efd9beed35fec21b4f2be14ad8453b4/src/library/methods/R/RMethodUtils.R#L1227-L1235).

> ```r
> .NonstandardGenericTest <- function(body, fname, stdBody)
> {
>     if(identical(body, stdBody))
>         FALSE
>     else if(.recursiveCallTest(body, fname))
>         TRUE
>     else
>         NA
> }
> ```

As far as I can tell, the initial version of this "nonStandardGenericFunction" was introduced in [revision 20995](https://github.com/r-devel/r-svn/commit/62c60532209aca868d5ec4560abcb49e53004f9d).

## Special case consequences

To see the consequences of triggering the special case or not, let's define two generics, which do the same thing, but one triggers the special case and the other does not:

```{r}
# This triggers the \pkg{methods} special case
setGeneric("myStandardGeneric", function(x) standardGeneric("myStandardGeneric"))

# This doesn't trigger the \pkg{methods} special case because of the braces
setGeneric("myCustomGeneric", function(x) {
  standardGeneric("myCustomGeneric")
})
```

And then let's define a simple method for both generics:

```{r}
setOldClass("rle")

setMethod("myStandardGeneric", "rle", function(x) {
  x$values * x$lengths / sum(x$lengths)
})

setMethod("myCustomGeneric", "rle", function(x) {
  x$values * x$lengths / sum(x$lengths)
})
```

### Performance

[What the methods package documentation seems to be suggesting](https://github.com/r-devel/r-svn/blob/56b976ae0de9cec972c8a90a8d301a9866094869/src/library/methods/man/Methods_Details.Rd#L282-L287), and what is often repeated, is that the standard generic shortcut may allow for better performance. We can try it on our simple example:

```{r}
x <- rpois(1000, lambda = 3) |> rle()
microbenchmark::microbenchmark(
  myStandardGeneric(x),
  myCustomGeneric(x),
  times = 1e6
)
```

This doesn't seem to have any measurable impact on performance.

On the topic of performance, [an email in the R-devel mailing list](https://www.mail-archive.com/r-devel@r-project.org/msg06307.html) says:

> standardGeneric() creates a  new context into which it copies 
bindings for all the objects in the frame of the generic function call

But my reading of this is that objects created before calling `standardGeneric()` in a non-standard generic that does some initial pre-computation would be copied into the context. It doesn't apply in the simple case of a non-standard generic that just wraps `standardGeneric()` in braces, as it doesn't do any pre-computation.

### Error stack

One GitHub comment mentions "an additional function call". If this the case, even if it doesn't have a measurable impact on performance, it could make the error stack more difficult to understand, and make debugging more difficult.

Let's try to trigger an error in both cases:

```{r, error = TRUE}
broken_rle <- structure(list(lengths = letters, values = 1:26), class = "rle")
myStandardGeneric(broken_rle)
myCustomGeneric(broken_rle)
```

The error stack appears to be the same in both cases, and it doesn't seem that the special case adds an extra function call.

### Lazy evaluation

[The `setGeneric()` documentation states](https://github.com/r-devel/r-svn/blob/56b976ae0de9cec972c8a90a8d301a9866094869/src/library/methods/man/setGeneric.Rd#L67-L70):

> A non-standard signature for the generic function may be
used to exclude arguments that take advantage of lazy evaluation;
in particular, if the argument may *not* be evaluated then it
cannot be part of the signature.



```{r}
setGeneric("foo", function(x = 1) standardGeneric("foo"))
setMethod("foo", signature(x = "missing"), function(x = 3) x)

setGeneric("foo_nonstandard", function(x = 1) { standardGeneric("foo_nonstandard") })
setMethod("foo_nonstandard", signature(x = "missing"), function(x = 13) x)

foo()
foo_nonstandard()
```

### Generic collision across packages

(Double check. But might a mechanism to avoid generic collision across packages be the original motivation for this special case? ?setGeneric() talks a bit about this in the "Specialized Local Generics" section.)

### Other issues

There might be other consequences of triggering the special case that I did not identify.
Some active Bioconductor package developers could not recall precisely **why** one should try to target the special case.

I did a [search on the Bioconductor mailing list](https://www.mail-archive.com/search?l=bioc-devel%40r-project.org&q=standardGeneric&start=100) and on the [Bioconductor slack/zulip] with the text "standardGeneric" and could not find any relevant threads.

## Summary



## Recommendations

### Advanced R book

It would be helpful to explain exactly **why** one should try to trigger the special case.
If this is not appropriate for the main text, perhaps an aside or footnote could be added.

The current phrasing ("more expensive") leads to believe the main issue is degraded performance, when this blog post seems to point in another direction.

### methods package

Upon exploring this issue, it appeared that non standard generics were originally meant to identify cases where some pre-computation is done in the generic rather than the methods, [which is discouraged](https://www.mail-archive.com/r-devel@r-project.org/msg06307.html):

> Maybe more relevant, if this question is motivated by a practical need, 
you may want to rethink what the generic function does.  "Nonstandard" 
generic functions are called that  for a reason.  The concept behind 
methods in S is that a generic function defines the _function_ or 
purpose of the computation, but it's only the methods that define how 
the purpose is carried out.
> 
> Using a nonstandard generic that does some extra computation always 
blurs the concept, and often makes methods harder to write or 
understand. 

But braced `standardGeneric()` calls do not actually perform any pre-computation, and it seems that it is purely a technical and historical quirk that they are treated as non-standard generics.

It would be helpful if braced `standardGeneric()` calls could be identified as standard generics.

There are several ways this could be achieved:

- By unbracing the generic body before the check in `.NonstandardGenericTest()`
- By comparing the body to both the braced and unbraced versions of the `standardGeneric()` call
- 
